A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2024-08-16 inst/hankin-AustrianJournalofStatistics.tex
work on issue #239

\documentclass{article}
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage{graphicx}%% uncomment this for including figures
\usepackage{setspace} \doublespacing
\usepackage{apacite}

@@

\title{A simplified Plackett--Luce likelihood function for rank orders}
%\author{Robin K. S. Hankin}
\maketitle

\begin{abstract}

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.
\end{abstract}

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley--Terry
  strengths] and say 20 independent complete rank observations,
\citeA{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

Consider a race between a competitor of Bradley--Terry strength $a$ and
$n$ cloned competitors, each of strength $b=1-a$.  An observation is
indexed by~$r$, the number of strength $b$ clones finishing ahead
of~$a$, and because we have $n+1$ competitors %%%%%%%%%%%%%%%%%%%%%%
$0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=n+(n-1)a$.  The Plackett--Luce likelihood
function~\cite{luce1959,plackett1975} for this situation would be

@@

One long-standing contention in the athletic world is that of lane
bias~\cite{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley--Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.474$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of Edwards's two units of
support criterion for rejection~\cite{edwards1972}.  This dataset
contains no evidence to support lane bias.

@@

Formula 1 motor racing is an important and prestigious motor sport
\cite{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.  However, as argued by
\citeA{hankin2023_formula1points}, the drivers' {\em ranks} are
statistically sufficient for analysis using Plackett--Luce approaches.
The methods presented here may be used to assess competitive strengths
of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over his
career from 2011 to 2023.  Checo's Formula 1 performance began
somewhat inauspiciously as second driver for Sauber for whom he
finished no higher than P7 in 2011, but his performance rose
subsequently, culminating in his achievement of a startling 20 podium
finishes in 2022--3.  Noting that Checo raced against 73 distinct
competitors (this rendering estimation of the other players'
Plackett--Luce strengths effectively impossible~\cite{hankin2020}), we
are nevertheless in a position to consider competitive strength using
the modified likelihood approach presented here.  We consider the
following inference problem:

@@

\bibliographystyle{apacite}
\bibliography{hyper2}    
\end{document}

>>>>

\documentclass[article]{ajs}
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage{graphicx}%% uncomment this for including figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Robin K. S. Hankin\,\orcidlink{0000-0001-5982-0415}\\
  University of Stirling\\ Scotland}
\title{A Simplified Plackett-Luce Likelihood Function for Rank Orders}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{A Simplified Plackett-Luce Likelihood Function for Rank Orders}

%% an abstract and keywords
\Abstract{

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.

}
\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
\Plainkeywords{keywords, comma-separated, non-capitalized,, R} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}
%% \setcounter{page}{1}
\Pages{1--xx}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  University of Stirling\\
  Scotland\\
  E-mail: \email{hankin.robin@gmail.com}\\
  URL: \url{https://www.stir.ac.uk/people/1966824}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% include your article here, just as usual

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley--Terry
  strengths] and say 20 independent complete rank observations,
\cite{hankin2017_rmd} shows that maximum likelihood strengths
identify the correct ranking with a probability of essentially zero.
The situation is worse in cases such as formula 1 motor racing
considered below, which requires us to consider ${\mathcal O}(100)$
competitors.

\cite{hankin2024_hyper3} considers Plackett-Luce likelihood functions
for a race between a competitor of Bradley--Terry strength $a$ and
$n=2$ or $n=3$ cloned competitors, each of strength $b=1-a$.  Here we
consider the generalized case with arbitrary $n$.  Under these
circumstances, an observation is indexed by~$r$, the number of
strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
initial field strength is $a+nb=n+(n-1)a$.  The Plackett--Luce
likelihood function~\citep{luce1959,plackett1975} for this situation
would be

@@

One long-standing contention in the athletic world is that of lane
bias~\citep{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Men's 100m Olympic running from
2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
competitors' ranks are used.  Figure~\ref{plotolymp} shows a
log-likelihood curve for the Bradley--Terry strength of the focal
competitor, here the runner in the innermost lane.  We see that the
evaluate of $0.474$ is very close to the neutral value of $0.5$, for
which the support is only $0.16$, far short of Edwards's two units of
support criterion for rejection~\citep{edwards1972}.  This dataset
contains no evidence to support lane bias.

@@

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.  However, as argued by
\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
statistically sufficient for analysis using Plackett--Luce approaches.
The methods presented here may be used to assess competitive strengths
of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over his
career from 2011 to 2023.  Checo's Formula 1 performance began
somewhat inauspiciously as second driver for Sauber for whom he
finished no higher than P7 in 2011, but his performance rose
subsequently, culminating in his achievement of a startling 20 podium
finishes in 2022--3.  Noting that Checo raced against 73 distinct
competitors---this rendering estimation of the other players'
Plackett--Luce strengths effectively
impossible~\citep{hankin2020}---we are nevertheless in a position to
consider competitive strength using the modified likelihood approach
presented here.  We consider the following inference problem:

@@

%\bibliographystyle{plainat}
\bibliography{hyper2}

\end{document}
